---
title: "ST558 Project1-Group K: Hanan Ali & Wenna Han"
format: html
editor: visual
---

## Data Processing

### First Steps - Read in one section of the data.

**1. Read in the data and select only few columns as required.**

```{r}
library(tidyverse)
data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/EDU01a.csv")
data <- data |>
  select(Area_name, STCOU, ends_with("D")) |>
  rename(area_name=Area_name)
data
```

**2. Convert the data into long format where each row has only one enrollment value for that Area_name.**

```{r}
library(tidyr)
data <- data |>
  pivot_longer(cols=3:12,
               names_to = "label",
               values_to = "enrollment_value")
data
```

**3. Parse the "label" string**\
For the "label" column, the first three characters represent the survey with the next four representing the type of value you have from that survey. The last two digits prior to the “D” represent the year of the measurement.

-   Parse the "label" string to pull out the year and convert the year into a numeric value such as 1997 or 2002.
-   Grab the first three characters and following four digits to create a new variable representing which measurement was grabbed.

```{r}
#create a function to convert two-digit character to four-digit numeric year.
convert_year <- function(year) {
  year <- as.numeric(year)
  ifelse(year > 24, 1900 + year, 2000 + year)
}

data <- data |>
  # parse the "label" string
  mutate(year=substr(label, 8, 9)) |>
  mutate(year=convert_year(year)) |>
  # grab measurement information
  mutate(measurement = substr(label, 1, 7))
data

```

**4. Create two data sets**

-   one data set that contains only non-county data
-   one data set that contains only county level data

```{r}
data_noncounty <- data [-grep(pattern = ", \\w\\w", data$area_name),]
data_noncounty
data_county <- data [grep(pattern = ", \\w\\w", data$area_name),]
data_county
```

-   For the county level data, add a class to the tibble called county.
-   Similarly, add a class to the non-county data called state.

```{r}
class(data_county) <- c("county", class(data_county))
class(data_county)
class(data_noncounty) <- c("state", class(data_noncounty))
class(data_noncounty)
```

**5. For the county level tibble, create a new variable that describes which state one of these county measurements corresponds to.**

```{r}
data_county <- data_county |>
  mutate(state = substr(area_name, nchar(area_name)-1, nchar(area_name)))
data_county
```

**6. For the non-county level tibble, create a new variable called “division” corresponding to the state’s classification of division here. If row corresponds to a non-state (i.e. UNITED STATES), return ERROR for the division. The code for this part will not be a ton of fun but can be made easier with the use of %in%.**

```{r}
# define the state’s classification of division
Northeast <- c("CONNECTICUT", "MAINE", "MASSACHUSETTS", "NEW HAMPSHIRE", 
               "NEW JERSEY", "NEW YORK", "PENNSYLVANIA", "RHODE ISLAND",
               "VERMONT")
Midwest <- c("ILLINOIS", "INDIANA", "IOWA", "KANSAS", "MICHIGAN", 
             "MINNESOTA", "MISSOURI", "NEBRASKA", "NORTH DAKOTA",
             "OHIO", "SOUTH DAKOTA", "WISCONSIN")
South <- c("ALABAMA", "ARKANSAS", "DELAWARE", "DISTRICT OF COLUMBIA",
           "FLORIDA", "GEORGIA", "KENTUCKY", "LOUISIANA", "MARYLAND",
           "MISSISSIPPI", "NORTH CAROLINA", "OKLAHOMA", "SOUTH CAROLINA",
           "TENNESSEE", "TEXAS", "VIRGINIA", "WEST VIRGINIA")
West <- c("ALASKA", "ARIZONA", "CALIFORNIA", "COLORADO", "HAWAII", "IDAHO",
          "MONTANA", "NEVADA", "NEW MEXICO", "OREGON", "UTAH", "WASHINGTON",
          "WYOMING")

data_noncounty <- data_noncounty |>
  # convert area_name to upper case
  mutate(area_name = toupper(area_name)) |>
  # classify the states
  mutate(division = ifelse(area_name == "UNITED STATES", "ERROR", 
                           ifelse (area_name %in% Northeast,
                                   "Northeast", 
                                   ifelse(area_name %in% Midwest,
                                          "Midwest", 
                                          ifelse(area_name %in% South,
                                                 "South",
                                                 ifelse(area_name %in% West, 
                                                        "West", 
                                                        #output CHECK if anything unexpected
                                                        "CHECK"))))))
data_noncounty
```

**Now we want to repeat the above process for the 2nd component of the data set.**

```{r}
data2 <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/EDU01b.csv")

#Crate a function for steps 1 and 2
function_for_step_1_2 <- function(data2, value_col_name = "enrollment_value") {
  data2 <- data2 |>
    select(Area_name, STCOU, ends_with("D")) |>
    rename(area_name = Area_name) |>
    pivot_longer(cols = 3:12, 
                 names_to = "label", 
                 values_to = value_col_name)
  return(data2)
}

#Process data
data2 <- function_for_step_1_2(data2)

#Crate a function for step 3
function_for_step_3 <- function(data2) {
  data2 <- data2 %>%
    mutate(year = as.numeric(substr(label, 8, 9)),
           year = ifelse(year > 24, 1900 + year, 2000 + year),
           measurement = substr(label, 1, 7))
  return(data2)
}

data2 <- function_for_step_3(data2)

function_for_step_5 <- function(data_county2) {
  data_county2 <- data_county2 %>%
    mutate(state = substr(area_name, nchar(area_name) - 1, nchar(area_name)))
  return(data_county2)
}

#Crate a function for step 6
function_for_step_6 <- function(data_noncounty2) {
  Northeast <- c("CONNECTICUT", "MAINE", "MASSACHUSETTS", "NEW HAMPSHIRE", 
                 "NEW JERSEY", "NEW YORK", "PENNSYLVANIA", "RHODE ISLAND",
                 "VERMONT")
  Midwest <- c("ILLINOIS", "INDIANA", "IOWA", "KANSAS", "MICHIGAN", 
               "MINNESOTA", "MISSOURI", "NEBRASKA", "NORTH DAKOTA",
               "OHIO", "SOUTH DAKOTA", "WISCONSIN")
  South <- c("ALABAMA", "ARKANSAS", "DELAWARE", "DISTRICT OF COLUMBIA",
             "FLORIDA", "GEORGIA", "KENTUCKY", "LOUISIANA", "MARYLAND",
             "MISSISSIPPI", "NORTH CAROLINA", "OKLAHOMA", "SOUTH CAROLINA",
             "TENNESSEE", "TEXAS", "VIRGINIA", "WEST VIRGINIA")
  West <- c("ALASKA", "ARIZONA", "CALIFORNIA", "COLORADO", "HAWAII", "IDAHO",
            "MONTANA", "NEVADA", "NEW MEXICO", "OREGON", "UTAH", "WASHINGTON",
            "WYOMING")
  
  data_noncounty2 <- data_noncounty2 %>%
    mutate(area_name = toupper(area_name),
           division = case_when(
             area_name == "UNITED STATES" ~ "ERROR",
             area_name %in% Northeast ~ "Northeast",
             area_name %in% Midwest ~ "Midwest",
             area_name %in% South ~ "South",
             area_name %in% West ~ "West",
             TRUE ~ "CHECK"
           ))
  return(data_noncounty2)
}

#Crate a function for final step
function_for_steps4_5_6 <- function(data2) {
  data_noncounty2 <- data2[-grep(pattern = ", \\w\\w", data2$area_name), ]
  data_county2 <- data2[grep(pattern = ", \\w\\w", data2$area_name), ]
  
  class(data_noncounty2) <- c("state", class(data_noncounty2))
  class(data_county2) <- c("county", class(data_county2))
  
  data_county2 <- function_for_step_5(data_county2)
  data_noncounty2 <- function_for_step_6(data_noncounty2)
  
  return(list(noncounty2 = data_noncounty2, county2 = data_county2))
}

#Process data
results <- function_for_steps4_5_6(data2)
data2_noncounty <- results$noncounty
data2_county <- results$county
```

**Creating wrapper function**

```{r}
# Define wrapper function
my_wrapper <- function(url, default_var_name = "enrollment_value") {
  data <- read_csv(url)
  
  # Function for steps 1 and 2
  data <- data %>%
    select(Area_name, STCOU, ends_with("D")) %>%
    rename(area_name = Area_name) %>%
    pivot_longer(cols = 3:12, names_to = "label", values_to = default_var_name)
  
  # Function for step 3
  data <- data %>%
    mutate(year = as.numeric(substr(label, 8, 9)),
           year = ifelse(year > 24, 1900 + year, 2000 + year),
           measurement = substr(label, 1, 7))
  
  # Function for steps 4, 5, and 6
  function_for_steps4_5_6 <- function(data2) {
    data_noncounty2 <- data2[-grep(pattern = ", \\w\\w", data2$area_name), ]
    data_county2 <- data2[grep(pattern = ", \\w\\w", data2$area_name), ]
    
    class(data_noncounty2) <- c("state", class(data_noncounty2))
    class(data_county2) <- c("county", class(data_county2))
    
    data_county2 <- function_for_step_5(data_county2)
    data_noncounty2 <- function_for_step_6(data_noncounty2)
    
    return(list(noncounty2 = data_noncounty2, county2 = data_county2))
  }
  
  # Process data
  results <- function_for_steps4_5_6(data)
  data_noncounty <- results$noncounty2
  data_county <- results$county2
  
  # Return the results
  return(list(noncounty = data_noncounty, county = data_county))
}

# Example usage:
url <- "https://www4.stat.ncsu.edu/~online/datasets/EDU01b.csv"
results <- my_wrapper(url)

# Access non-county and county data
data_noncounty3 <- results$noncounty
data_county3 <- results$county
```

**Call it and combine your data**

```{r}
# URLs for the two CSV files
url1 <- "https://www4.stat.ncsu.edu/~online/datasets/EDU01a.csv"
url2 <- "https://www4.stat.ncsu.edu/~online/datasets/EDU01b.csv"  

# Call the wrapper function for the first dataset
results1 <- my_wrapper(url1, default_var_name = "enrollment_value")

# Process data
data1_noncounty <- results1$noncounty
data1_county <- results1$county

# Call the wrapper function for the second dataset
results2 <- my_wrapper(url2, default_var_name = "enrollment_value")

# Process data
data2_noncounty <- results2$noncounty
data2_county <- results2$county
```

**Single short function**

```{r}
#Creat combined short function
combine <- function(results1, results2) {
  combined_county <- bind_rows(results1$county, results2$county)
  combined_noncounty <- bind_rows(results1$noncounty, results2$noncounty)
  return(list(county = combined_county, noncounty = combined_noncounty))
}

#Process data
combined_results <- combine(results1, results2)
combined_county <- combined_results$county
combined_noncounty <- combined_results$noncounty
```

### Writing a Generic Function for Summarizing

**write a function that plots the mean value of the statistic (enrollment for this data set) across the years for each Division. That is, on the x-axis we want the numeric year value, on the y-axis we want the mean of the statistic for each Division and numeric year. Also, we want to remove observations at the ERROR setting of Division.**

```{r}
plot.state <- function(df, var_name = "_your_default_value_"){
  # Calculate the mean of the statistic for each Division and year
  df_summary <- df |>
    filter(division != "ERROR") |>  # Remove the ERROR setting
    group_by(year, division) |>  # Group by year and division
    summarize(mean_statistic = mean(get(var_name)))  # Calculate mean
  
  # Create a line plot with ggplot
  ggplot(df_summary, aes(x = year, y = mean_statistic, color = division)) +
    geom_line() +
    labs(title = paste("Mean", var_name, "across Years by Division"),
         x = "year",
         y = paste("Mean", var_name),
         color = "division")
}
```

```{r, eval=FALSE, echo=FALSE}
# test out this function
plot.state(combined_noncounty, var_name = "enrollment_value")
```

**For the class county we’ll do a similar plotting function but with more flexibility. This function should allow the user to:**\
• specify the state of interest, giving a default value if not specified.\
• determine whether the ‘top’ or ‘bottom’ most counties should be looked at with a default for ‘top’.\
• instruct how many of the ‘top’ or ‘bottom’ will be investigated with a default value of 5

```{r}
plot.county <- function(df, State = "AL", 
                        var_name = "enrollment_value", 
                        top_bottom = "top", 
                        num_counties = 5) {
  # Filter data for the specified state
  state_data <- df |>
    filter(state == State)
  
  county_means <- state_data |>
    group_by(area_name) |> # Group by area_name
    summarize(mean_statistic = mean(get(var_name))) |> # Calculate mean
    arrange(mean_statistic)
    
  # Sort values based on 'top' or 'bottom' specification
  if (top_bottom == "top") {
    sorted_counties <- county_means |>
      arrange(desc(mean_statistic)) |>
      head(num_counties)
  } else if (top_bottom == "bottom") {
    sorted_counties <- county_means |>
      arrange(mean_statistic) |>
      head(num_counties)
  } else {
    stop("Invalid value for top_bottom")
  }

  # Filter the data for the specified counties
  plot_data <- state_data |>
    filter(area_name %in% sorted_counties$area_name)
  
  # Create the plot
  ggplot(plot_data, aes(x = year, y = get(var_name), color = area_name)) +
    geom_line() +
    labs(title = paste("Mean", var_name, "for", top_bottom, num_counties, "Counties in", State),
      x = "Year",
      y = paste("Mean", var_name),
      color = "County")
}
```

```{r, eval=FALSE, echo=FALSE}
# test out this function
plot.county(combined_county, State = "AK")
plot.county(combined_county, State = "AZ", top_bottom = "bottom", num_counties = 7)
```

**Put it together section**

```{r}
# URLs for the two CSV files given previously
url1 <- "https://www4.stat.ncsu.edu/~online/datasets/EDU01a.csv"
url2 <- "https://www4.stat.ncsu.edu/~online/datasets/EDU01b.csv"  

# Process data - URL1
results1 <- my_wrapper(url1, default_var_name = "enrollment_value")
data1_noncounty <- results1$noncounty
data1_county <- results1$county
# Process data - URL2
results2 <- my_wrapper(url2, default_var_name = "enrollment_value")
data2_noncounty <- results2$noncounty
data2_county <- results2$county

# Run your data combining function
combined_results <- combine(results1, results2)
combined_county <- combined_results$county
combined_noncounty <- combined_results$noncounty

# the plot function on the state data frame
plot.state(combined_noncounty, var_name = "enrollment_value")

# Use the plot function on the county data frame
# Once specifying the state to be “NC”, the group being the top, the number looked at being 20
plot.county(combined_county, State = "NC", top_bottom = "top", num_counties = 20)

# Once specifying the state to be “SC”, the group being the bottom, the number looked at being 7
plot.county(combined_county, State = "SC", top_bottom = "bottom", num_counties = 7)

# Once without specifying anything (defaults used)
plot.county(combined_county)

# Once specifying the state to be “PA”, the group being the top, the number looked at being 8
plot.county(combined_county, State = "PA", top_bottom = "top", num_counties = 8)
```

**Lastly, read in another couple similar data sets and apply your functions!**

```{r}
# URLs for the two CSV files given previously
url3 <- "https://www4.stat.ncsu.edu/~online/datasets/PST01a.csv"
url4 <- "https://www4.stat.ncsu.edu/~online/datasets/PST01b.csv"  
url5 <- "https://www4.stat.ncsu.edu/~online/datasets/PST01c.csv"
url6 <- "https://www4.stat.ncsu.edu/~online/datasets/PST01d.csv"

# Process data - URL3
results3 <- my_wrapper(url3, default_var_name = "enrollment_value")
data3_noncounty <- results3$noncounty
data3_county <- results3$county
# Process data - URL4
results4 <- my_wrapper(url4, default_var_name = "enrollment_value")
data4_noncounty <- results4$noncounty
data4_county <- results4$county
# Process data - URL5
results5 <- my_wrapper(url5, default_var_name = "enrollment_value")
data5_noncounty <- results5$noncounty
data5_county <- results5$county
# Process data - URL6
results6 <- my_wrapper(url6, default_var_name = "enrollment_value")
data6_noncounty <- results6$noncounty
data6_county <- results6$county

#Run your data combining function
combined_results34 <- combine(results3, results4)
combined_results345 <- combine(combined_results34, results5)
combined_results3456 <- combine(combined_results345, results6)
combined_county3456 <- combined_results3456$county
combined_noncounty3456 <- combined_results3456$noncounty

# Use the plot function on the state data frame
plot.state(combined_noncounty3456, var_name = "enrollment_value")

# Use the plot function on the county data frame
# Once specifying the state to be “CA”, the group being the top, the number looked at being 15
plot.county(combined_county3456, State = "CA", top_bottom = "top", num_counties = 15)

# Once specifying the state to be “TX”, the group being the top, the number looked at being 4
plot.county(combined_county3456, State = "TX", top_bottom = "top", num_counties = 4)

#Once without specifying anything (defaults used)
plot.county(combined_county3456)

# Once specifying the state to be “NY”, the group being the top, the number looked at being 10
plot.county(combined_county3456, State = "NY", top_bottom = "top", num_counties = 10)
```
